---
title: "assignment_1_extra_code"
author: "Manuel Thomasen"
date: "2022-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
average_mlu <- log(1.5)
sd_mlu_asd <- log(1.5)-log(1.5-0.5)
sd_mlu_td <- log(1.5)-log(1.5-0.3)
change_mlu_asd <- 0.4
change_mlu_td <- 0.6
change_sd_mlu_asd <- 0.4
change_sd_mlu_td <- 0.2

sample_size <- 100
```



doesn't work, because if can only look at one variable at a time and does not repeat the process after that one variable 
- Thange if to a ifelse function
- chnage rnorm to rlnorm (hmmm, maybe not)
```{r}
data_sim_fun <- function(n, mlu_mean, mlu_sd_td, mlu_sd_asd, b_mean_td, b_mean_asd, b_sd_td, b_sd_asd){
  df <- tibble(ID=seq(n+n),
                  diagnosis=rep(c('TD', 'ASD'), each=n),
                  if (diagnosis=='TD'){
                    mutate(
                      MLU=rnorm(
                        n,mean=mlu_mean,sd=mlu_sd_td),
                      Beta=rnorm(
                        n, mean=b_mean_td, sd=b_sd_td),
                      error=0.2)
                    }
                  else{
                    mutate(
                      MLU=rnorm(
                        n,mean=mlu_mean,sd=mlu_sd_asd),
                      Beta=rnorm(
                        n, mean=b_mean_asd, sd=b_sd_asd),
                      error=0.2)
                    }
                  )
                  
  return(df)
}

df <- data_sim_fun(sample_size, average_mlu, sd_mlu_td, sd_mlu_asd, change_mlu_td, change_mlu_asd, change_sd_mlu_td, change_sd_mlu_asd)


```


works, but not very pretty
```{r}
data_sim_asd <- function(n, mlu_mean, mlu_sd, b_mean, b_sd){
 td_df <- tibble(ID=(seq(n), 
                  MLU=rnorm(n, mean=mlu_mean,sd=mlu_sd),
                  Beta=rnorm(n, mean=b_mean, sd=b_sd),
                  error=0.2)
  return(td_df)
}

td_df <- data_sim_td(sample_size, average_mlu, sd_mlu_td, change_mlu_td, change_sd_mlu_td)

data_sim_asd <- function(n, mlu_mean, mlu_sd, b_mean, b_sd){
 asd_df <- tibble(ID=(seq(n)+n), 
                  MLU=rnorm(n, mean=mlu_mean,sd=mlu_sd),
                  Beta=rnorm(n, mean=b_mean, sd=b_sd),
                  error=0.2)
  return(asd_df)
}

asd_df <- data_sim_asd(sample_size, average_mlu, sd_mlu_asd, change_mlu_asd, change_sd_mlu_asd)

expand_td <- tibble(
  expand.grid(
    ID=seq(sample_size),
    visit=seq(6)
  )
)

td_df <- merge(td_df, expand_td)

expand_asd <- tibble(
  expand.grid(
    ID=(seq(sample_size)+sample_size),
    visit=seq(6)
  )
)

asd_df <- merge(asd_df, expand_asd)

td_df$condition <- 'TD'

asd_df$condition <- 'ASD'

sim_df <- rbind(asd_df, td_df)

for (i in seq(nrow(sim_df))){
  sim_df$MLU_true[i] = exp(rnorm(1, sim_df$MLU[i] + sim_df$Beta[i] * (sim_df$visit[i]-1), sim_df$error[i]))
}
```



Simulation work but get one beta value for each subject and does not vary between visits 
```{r}
data_sim_fun <- function(n, mlu_mean, mlu_sd_td, mlu_sd_asd, b_mean_td, b_mean_asd, b_sd_td, b_sd_asd){
  df_1 <- tibble(ID=seq(n+n),
                  diagnosis=rep(c('TD', 'ASD'), each=n))
  df_1$MLU_int <- ifelse(df_1$diagnosis=='TD',
                   rnorm(n,mean=mlu_mean,sd=mlu_sd_td),
                   rnorm(n,mean=mlu_mean,sd=mlu_sd_asd))
  df_1$beta <- ifelse(df_1$diagnosis=='TD', 
                    rnorm(n, mean = b_mean_td, sd=mlu_sd_td), 
                    rnorm(n, mean=b_mean_asd, sd=b_sd_asd))
  df_1$error <- 0.2
  expand_df <- tibble(
    expand.grid(
      ID=seq(n+n),
      visit=seq(6)))
  df <- merge(df_1, expand_df)
  for (i in seq(nrow(df))){
    df$MLU_true[i] = exp(rnorm(1, df$MLU_int[i] + df$beta[i] * (df$visit[i]-1), df$error[i]))
}
  return(df)
}

df <- data_sim_fun(sample_size, average_mlu, sd_mlu_td, sd_mlu_asd, change_mlu_td, change_mlu_asd, change_sd_mlu_td, change_sd_mlu_asd)
```
```{r}
int_asd <- rlnorm(n, mean=average_mlu, sd=sd_mlu_asd)
int_td <- rlnorm(n, mean=average_mlu, sd=sd_mlu_td)

slope_asd <- rnorm(n, mean=change_mlu_asd, sd=change_sd_mlu_asd)
slope_td <- rnorm(n, mean = change_mlu_td, sd=change_sd_mlu_td)

error_td <- rlnorm(n, mean=0, sd=e)
error_asd <- rlnorm(n, mean=0, sd=e)
```

sim_data <- 
  tibble(diagnosis=rep(c('TD', 'ASD'), each=n)) %>% 
  mutate(intercept=ifelse(diagnosis=='TD', int_td, int_asd)) %>% 
  mutate(slope=ifelse(diagnosis=='TD', slope_td, slope_asd)) %>% 
  mutate(error=ifelse(diagnosis=='TD', error_td, error_asd)) %>% 
  dplyr::mutate(ID=row_number()) %>% 
  slice(rep(1:n(), each=6)) %>% 
  add_column(visit=rep(c(1,2,3,4,5,6), times=n+n))

for(i in seq(nrow(sim_data))){
  sim_data$MLU[i]=exp(rnorm(1, sim_data$intercept[i]+sim_data$slope[i]*(sim_data$visit[i]-1), sim_data$error[i]))
}

int_asd <- rlnorm(sample_size, mean=average_mlu, sd=sd_mlu_asd)
int_td <- rlnorm(sample_size, mean=average_mlu, sd=sd_mlu_td)

slope_asd <- rnorm(sample_size, mean=change_mlu_asd, sd=change_sd_mlu_asd)
slope_td <- rnorm(sample_size, mean = change_mlu_td, sd=change_sd_mlu_td)

error_td <- rlnorm(sample_size, mean=0, sd=e)
error_asd <- rlnorm(sample_size, mean=0, sd=e)

data_sim <- 
  tibble(diagnosis=rep(c('TD', 'ASD'), each=n))
data_sim$MLU_int <-
  ifelse(data_sim$diagnosis=='TD',int_td,int_asd)
data_sim$beta <- ifelse(data_sim$diagnosis=='TD', slope_td, slope_asd)
data_sim$error <- ifelse(data_sim$diagnosis=='TD', error_td, error_asd)
data_sim <- dplyr:: mutate(ID=row_number()) %>% slice(rep(1:n()), each=6)
  
  expand_df <- tibble(
    expand.grid(
      ID=seq(sample_size+sample_size),
      visit=seq(6)))
  df <- merge(data_sim, expand_df)

  for (i in seq(nrow(df))){
    df$MLU[i] = exp(rnorm(1, df$intercept[i] + df$slope[i] * (df$visit[i]-1), df$error[i]))
}
  

df <- data_sim_fun(sample_size, average_mlu, sd_mlu_td, sd_mlu_asd, change_mlu_td, change_mlu_asd, change_sd_mlu_td, change_sd_mlu_asd)
